A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-02-25 analysis/paper/paper.Rmd
why do I still see t-test in paper.docx?

# Introduction {-}

<!---network analysis approaches and basic concepts-->
Network analysis has been increasingly used by archaeologists to understand past relationships, interactions, and structures of observed phenomena by visualizing and analyzing rational data [@Brandes2013; @Mills2017; @Peeples2019]. The recent development of modeling approaches enable hypothesis testing for the intercorrelations between individual elements and overall structures inherent in networks [@Brughmans2013; @Brughmans2018; @Freeman2004; @Salvini2010]. By comparing hypothesized networks with the observed network, network modeling can answer anthropological questions related to exchange [@Gjesfjeld2013; @Crabtree2015], diffusion [@Ostborn2015], or social transformation [@Mills2013] in a statistical way. Among network modeling methods, exponential random graph models (ERGMs) are stochastic models for investigating the process of network formation through dependence assumptions for relationships and simulations for network patterns [@Harris2013; @Handcock2008; @Ghafouri2020]. Since relationships in a network emerging relative to their position in a network can be hypothesized, ERGMs are promising to evaluate dynamic social processes behind observed archaeological networks [@Brughmans2014]. However, computational difficulties and the sensitivity to uncertainties to some extent limits the practical applications of ERGMs, where results could be challenging for interpretation [@Caimo2014].

<!--study social changes in pericolonial context using burial data as a proxy-->
In this paper, we go beyond ERGMs with a novel Bayesian approach that can alleviate the computational issues in ERGMs to better interpret the process of network formation for past phenomena. Using burial data from an Iron Age site in northeastern Taiwan as a case study, we explore social changes by investigating the formation of material connections between burials. Social changes in Indigenous societies, when faced with colonial powers, are commonly observed in many parts of the world, especially European colonies where Indigenous economic, cultural, or socio-political aspects were substantially impacted [@Dietler2005; @Silliman2005; @Voss2005]. Recent studies demonstrate that the indirect effect of colonialism, or involvement in long-distance trade, could also have impacts on Indigenous societies, this is known as a pericolonial context [@Acabado2017; @Trabert2017; @LWandBM2020ornament]. To explore social changes, burials are important evidence to understand past societies because material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased, and social relations between them [@Binford1971; @Drennan2010; @Saxe1970]. Burial treatments can further indicate social complexity or inequality where foreign goods from long-distance trade were used by Indigenous people to express status [@Carter2015; @Dolfini2019].

<!--Our burial network and research implication-->
Usually, burial studies are based on characterizing each physical trace or burial variable, such as biological records, grave forms and goods, or ritual behaviors, and finally combining those individual observations to infer past societies [@Byrd1995; @Seikel2011]. Here we treat burials as a complex network based on the assumption that a network is a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data [@Brandes2013]. Our case study illustrates how a Bayesian approach to network modeling provides new insights for studying the formation of archaeological networks and allows testing on theoretical assumptions to better understand dynamic social processes.

<!-- intro to networks: move to SI -->
<!-- two types of network analysis: move to SI -->

# Exponential random graph models in a Bayesian framework {-}

<!-- ERGMs in archaeology--> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. By fitting models with the observed network of archaeological data, they proposed that ERGMs is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively.

<!--limitations of ERGMs-->
However, those studies also point out some limitations of ERGMs, such as sensitivity to missing data and less able to handle uncertainty. Also, it is difficult in ERGMs to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with a larger set of networks [@Caimo2017bayesian]. This is also termed “doubly intractable” since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue where probability models tend to overestimate a small number of extreme graphs by assigning too much weight, e.g., in extreme cases such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs by means of a Bayesian framework.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations (details in Supplementary Online Materials) to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. With the exchange algorithm, Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives better convergence results. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with the approximate exchange algorithm, a Bayesian approach generates posterior probabilities that incorporate our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by examining the posterior mean and 95% credible intervals [@Caimo2017bayesian]. In addition, Bayesian approaches are useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on average, 80% of the ties when a third of data is missing.

# Background {-}

## A Case Study from Northeastern Taiwan {-}

<!--archaeological specific case study and our model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. Our general model proposes that the influence of a colonial power, combined with high local values attached to imported goods, leads to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. Increased social inequality associated with the use of foreign prestige goods is recognized as network strategies for gaining power [@Blanton1996; @Feinman2000]. Network-based societies are identified in some places in Asia with involvement in long-distance trade [@Carter2015; @Ueda2016; @Liu2006]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

<!--theoretical framework and research questions-->
The features of network-based societies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks [@Blanton1996; @Feinman2000], while corporate-based societies stress communal ritual elements, shared power, and wealth differentiation, if any, would be associated with corporate groups, such as age groups [@Siegel1999]. Using the theoretical framework of corporate-network strategies, or horizontal hierarchy, we test a hypothesis that changes from a corporate mode (a structure showing more subgroups with less wealth differentiation ) to a network mode (a centralized structure with more wealth differentiation) can be observed in the Kiwulan burial network after European arrival. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) if so, what are the major variables affecting or forming unequal social positions that might hint at social heterogeneity? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

## Materials {-}

@@

<!-- Kiwulan burials and a review of current studies of social organization -->
We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) [@Chen2007]. A total of 90 burials were unearthed from adjacent squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). The preservation of human remains at Kiwulan was generally poor, so we have incomplete age and sex data (details in Supplementary Online Materials) and lack health status data. Burials are oriented in an east-west direction on the north side of the residential area, which is indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement of houses. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on the comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing the phases before and after European arrival. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until arrival of Han Chinese immigrants).

@@

<!-- Our chronology and principles of network building -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. We also excluded `r disturb` burials that were heavily disturbed by modern construction because we cannot determine their chronology.

Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, and fish-shaped ornaments. These items are considered as high-value across different archaeological contexts based on their rarity and descriptions in historical records [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads’ high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as having higher status in their community [@LiandWu2006].

# Methods {-}

@@

## Hypothesis and construction of networks {-}
<!-- our expectation based on hypothesis --> 
Our pericolonial impact hypothesis is that differential access to foreign prestige goods after the European presence led to increased social inequality at Kiwulan, where social structure changed from a more corporate strategy to a more network strategy [@Blanton1996; @Drennan2010; @Feinman2000]. This would be reflected by the pattern of intercorrelations of foreign prestige goods in burial networks. If social inequality gradually increased as we hypothesize, then we expect to observe a network structure with higher centralization (or popularity) and less transitivity (or clustering), along with associations with wealth accumulation, indicative of a network-based society. In contrast, the network before the European arrival will show low centralization and high transitivity, associated with ritual element, or age/sex groups to indicate a corporate-based society. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- burial data and the definition of tie formation  --> 
Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they both possess each type of goods (i.e. presence or absence). Node attributes here include osteologically determined age and sex, cultural data, such as ritual pottery, and a burial value index. Ritual pottery was identified as locally made ceramics placed above an individual. Those pots mostly have soot on vessel exteriors and charred residues on interiors that are interpreted as vessels for funeral feasting [@Hsieh2009]. We calculated our burial value by summing values of all types of prestige goods in each burial context. For example, for gold-foil beads, we take the total number of burials (90) and divide by number of burials with gold-foil beads (46), to give a value of 90/46 = 1.96. We repeat this for each type of prestige item. For each burial we sum these values of each type of prestige item to get a burial value [@Jorgensen1991].

We then assigned the burials into three ranks according to breaks in the distribution of burial values, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth (details in Supplementary Online Materials). Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

@@

## Model specification in a Bayesian framework {-}

<!-- Network construction and specification -->
We quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014]. Table \@ref(tab:table-ergm-terms) lists the network parameters we used for dependence assumptions that define our models [@Morris2008]. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of prestige goods between two burials. Based on our hypothesis, we model a network with increased social inequality to be represented by endogenous network effects: low transitivity and high centralization. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to explore the importance of these variables in tie forming. For example, if age-homophily is important here, then people of the same age should have the same burial goods. We also include the physical distance between burials as an indicator of a kinship-based relations since the deceased from the same family were buried nearby [@LiandWu2006]. Our model may reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others, with more wealth differentiation after the European arrival.

```{r table-ergm-terms}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
ergm_terms_selected <-
  ergm_terms %>% 
  select(-"Description", -"note")
knitr::kable(ergm_terms_selected,
             caption = "Network parameters in ERGMs used for model specifications with associated archaeological interpretation for burial relations. ERGM terminology is from the R package statnet (Handcock, 2008; Morris, 2008)")
```

@@

## Choice of prior values to evaluate our anthropological model {-}

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distributions for the priors are typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of the edge density parameter to low for both network models, which follows a normal distribution with mean at `r priors$mean_same[1]`, and standard deviations at `r priors$sigma_same[1]` (i.e (N(`r priors$mean_same[1]`, `r priors$sigma_same[1]`))). For covariates about biological features, such as age, sex, we specified a vague prior (N(`r priors$mean_same[2]`, `r priors$sigma_same[2]`)) for both models to explore their effects. For physical distance between burials, we also set a vague prior effect (N(`r priors$mean_same[8]`, `r priors$sigma_same[8]`)) to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances. To evaluate our anthropological model of increased social inequality over time based on assumptions of corporate-network strategies [@Drennan2010; @Feinman2000], we incorporated different prior information for the network variables that are meaningful for social inequality, especially for transitivity and centralization. We set the priors to higher transitivity (N(`r priors$pre_prior_mean[6]`, `r priors$pre_prior_sigma[6]`)), lower centralization (N(`r priors$pre_prior_mean[7]`, `r priors$pre_prior_sigma[7]`)), and higher covariate effect of ritual activity (N(`r priors$pre_prior_mean[4]`, `r priors$pre_prior_sigma[4]`)) for the network before European contact to indicate less social differentiation and stress ritual element shared in corporate groups. Conversely, we set the priors for the network after European arrival to lower transitivity (N(`r priors$post_prior_mean[6]`, `r priors$post_prior_sigma[6]`)), higher centralization (N(`r priors$post_prior_mean[7]`, `r priors$post_prior_sigma[7]`)), and higher covariate effect of burial values (N(`r priors$post_prior_mean[5]`, `r priors$post_prior_sigma[5]`)) to model an increased social differentiation.

@@

# Reproducibility and open source materials {-}

The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://doi.org/xxx/xxx to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

# Results {-}

>>>>

# Introduction {.unnumbered}

<!---network analysis approaches and basic concepts-->

Network analysis has been increasingly used by archaeologists to understand past relationships, interactions, and structures of observed phenomena by visualizing and analyzing rational data [@Brandes2013; @Mills2017; @Peeples2019]. The recent development of modeling approaches enable hypothesis testing for the intercorrelations between individual elements and overall structures inherent in networks [@Brughmans2013; @Brughmans2018; @Freeman2004; @Salvini2010]. By comparing hypothesized networks with the observed network, network modeling can answer anthropological questions related to exchange [@Gjesfjeld2013; @Crabtree2015], diffusion [@Ostborn2015], or social transformation [@Mills2013] in a statistical way. Among network modeling methods, exponential random graph models (ERGMs) are stochastic models for investigating the process of network formation through dependence assumptions for relationships and simulations for network patterns [@Harris2013; @Ghafouri2020]. Since relationships in a network emerging relative to their position in a network can be hypothesized, ERGMs are promising to evaluate dynamic social processes behind observed archaeological networks [@Brughmans2014]. However, computational difficulties and the sensitivity to uncertainties to some extent limits the practical applications of ERGMs, where results could be challenging for interpretation [@Caimo2014].

<!--study social changes in pericolonial context using burial data as a proxy-->

In this paper, we go beyond ERGMs with a novel Bayesian approach that can alleviate the computational issues in ERGMs to better interpret the process of network formation for past phenomena. Using burial data from an Iron Age site in northeastern Taiwan as a case study, we explore social changes by investigating the formation of material connections between burials. Social changes in Indigenous societies, when faced with colonial powers, are commonly observed in many parts of the world, especially European colonies where Indigenous economic, cultural, or socio-political aspects were substantially impacted [@Dietler2005; @Silliman2005; @Voss2005]. Recent studies demonstrate that the indirect effect of colonialism, or involvement in long-distance trade, could also have impacts on Indigenous societies, this is known as a pericolonial context [@Acabado2017; @Trabert2017; @LWandBM2020ornament]. To explore social changes, burials are important evidence to understand past societies because material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased, and social relations between them [@Binford1971; @Drennan2010; @Saxe1970]. Burial treatments can further indicate social complexity or inequality where foreign goods from long-distance trade were used by Indigenous people to express status [@Carter2015; @Dolfini2019].

<!--Our burial network and research implication-->

Usually, burial studies are based on characterizing each physical trace or burial variable, such as biological records, grave forms and goods, or ritual behaviors, and finally combining those individual observations to infer past societies [@Byrd1995; @Seikel2011]. Here we treat burials as a complex network based on the assumption that a network is a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data [@Brandes2013]. Our case study illustrates how a Bayesian approach to network modeling provides new insights for studying the formation of archaeological networks and allows testing on theoretical assumptions to better understand dynamic social processes.

<!-- intro to networks: move to SI -->

<!-- two types of network analysis: move to SI -->

# Exponential random graph models in a Bayesian framework {.unnumbered}

<!-- ERGMs in archaeology-->

Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. By fitting models with the observed network of archaeological data, they proposed that ERGMs is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively.

<!--limitations of ERGMs-->

However, those studies also point out some limitations of ERGMs, such as sensitivity to missing data and less able to handle uncertainty. Also, it is difficult in ERGMs to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with a larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue where probability models tend to overestimate a small number of extreme graphs by assigning too much weight, e.g., in extreme cases such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs by means of a Bayesian framework.

@@

<!-- Bayesian inference for Exponential Random Graph Models -->

Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations (details in Supplementary Online Materials) to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. With the exchange algorithm, Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives better convergence results. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with the approximate exchange algorithm, a Bayesian approach generates posterior probabilities that incorporate our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by examining the posterior mean and 95% credible intervals [@Caimo2017bayesian]. In addition, Bayesian approaches are useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on average, 80% of the ties when a third of data is missing.

# Background {.unnumbered}

## A Case Study from Northeastern Taiwan {.unnumbered}

<!--archaeological specific case study and our model-->

We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. Our general model proposes that the influence of a colonial power, combined with high local values attached to imported goods, leads to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. Increased social inequality associated with the use of foreign prestige goods is recognized as network strategies for gaining power [@Blanton1996; @Feinman2000]. Network-based societies are identified in some places in Asia with involvment in long-distance trade [@Carter2015; @Ueda2016; @Liu2006]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

<!--theoretical framework and research questions-->

The features of network-based societies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks [@Blanton1996; @Feinman2000], while corporate-based societies stress communal ritual elements, shared power, and wealth differentiation, if any, would be associated with corporate groups, such as age groups [@Siegel1999]. Using the theoretical framework of corporate-network strategies, we test a hypothesis that changes from a corporate mode (a structure showing more subgroups with less wealth differentiation ) to a network mode (a centralized structure with more wealth differentiation) can be observed in the Kiwulan burial network after European arrival. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) if so, what are the major variables affecting or forming unequal social positions that might hint at social heterogeneity? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

## Materials {.unnumbered}

@@

<!-- Kiwulan burials and a review of current studies of social organization -->

We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) [@Chen2007]. A total of 90 burials were unearthed from adjacent squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, which is indicated by post-holes and *in-situ* wooden posts, suggesting a well-organized spatial arrangement of houses. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on the comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing the phases before and after European arrival. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until arrival of Han Chinese immigrants).

@@

<!-- Our chronology and principles of network building -->

To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. We also excluded `r disturb` burials that were heavily disturbed by modern construction because we cannot determine their chronology. Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, and fish-shaped ornaments. These items are considered as high-value across different archaeological contexts based on their rarity and descriptions in historical records [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads' high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as having higher status in their community [@LiandWu2006].

# Methods {.unnumbered}

@@

## Hypothesis and construction of networks {.unnumbered}

<!-- our expectation based on hypothesis -->

Our pericolonial impact hypothesis is that differential access to foreign prestige goods after the European presence led to an increased social inequality at Kiwulan, where social structure changed from a more corporate strategy to a more network strategy [@Blanton1996; @Drennan2010; @Feinman2000]. This would be reflected by the pattern of intercorrelations of foreign prestige goods in burial networks. If social inequality gradually increased as we hypothesize, then we expect to observe a network structure with higher centralization (or popularity) and less transitivity (or clustering), along with associations with wealth accumulation, indicative of a network-based society. In contrast, the network before the European arrival will show low centralization and high transitivity, associated with ritual element, or age/sex groups to indicate a corporate-based society. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- burial data and the definition of tie formation  -->

Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they both possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and a burial value index. Ritual pottery was identified as locally made ceramics placed above an individual. Those pots mostly have soot on vessel exteriors and charred residues on interiors that are interpreted as vessels for funeral feasting [@Hsieh2009]. We calculated our burial value by summing values of all types of prestige goods in each burial context. The value of prestige good is calculated by taking the total number of burials in Kiwulan, and then dividing this by the number of burials with each type of prestige good [@Jorgensen1991]. We then assigned the burials into three ranks according to their burial values, high (\>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (\<`r ave_value`), as an index of wealth (Details about other node attributes in Supplementary Online Materials). Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

@@

## Model specification in a Bayesian framework {.unnumbered}

<!-- Network construction and specification -->

We quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014]. Table \@ref(tab:table-ergm-terms) lists the network parameters we used for dependence assumptions that define our models [@Morris2008]. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of prestige goods between two burials. Based on our hypothesis, we model a network with increased social inequality to be represented by endogenous network effects: low transitivity and high centralization. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to explore the importance of these variables in tie forming. For example, if age-homophily is important here, then people of the same age should have the same burial goods. We also include the physical distance between burials as an indicator of a kinship-based relations since the deceased from the same family were buried nearby [@LiandWu2006]. Our model may reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others, with more wealth differentiation after the European arrival.

```{r table-ergm-terms}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
ergm_terms_selected <-
  ergm_terms %>% 
  select(-"Description", -"note")
knitr::kable(ergm_terms_selected,
             caption = "Network parameters in ERGMs used for model specifications with associated archaeological interpretation for burial relations.")
```

@@

## Choice of prior values to evaluate our anthropological model {.unnumbered}

<!-- add text about prior setting and anthropological assumptions of our prior values -->

Normal distributions for the priors are typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of the edge density parameter to low for both network models, which follows a normal distribution with mean at `r priors$mean_same[1]`, and standard deviations at `r priors$sigma_same[1]` (i.e (N(`r priors$mean_same[1]`, `r priors$sigma_same[1]`))). For covariates about biological features, such as age, sex, we specified a vague prior (N(`r priors$mean_same[2]`, `r priors$sigma_same[2]`)) for both models to explore their effects. For physical distance between burials, we also set a vague prior effect (N(`r priors$mean_same[8]`, `r priors$sigma_same[8]`)) to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances. To evaluate our anthropological model of increased social inequality over time based on assumptions of corporate-network strategies [@Drennan2010; @Feinman2000], we incorporated different prior information for the network variables that are meaningful for social inequality, especially for transitivity (gwesp) and centralization (gwedegree). We set the priors to higher transitivity (N(`r priors$pre_prior_mean[6]`, `r priors$pre_prior_sigma[6]`)), lower centralization (N(`r priors$pre_prior_mean[7]`, `r priors$pre_prior_sigma[7]`)), and higher covariate effect of ritual activity (N(`r priors$pre_prior_mean[4]`, `r priors$pre_prior_sigma[4]`)) for the network before European contact to indicate less social differentiation and stress ritual element shared in corporate groups. Conversely, we set the priors for the network after European arrival to lower transitivity (N(`r priors$post_prior_mean[6]`, `r priors$post_prior_sigma[6]`)), higher centralization (N(`r priors$post_prior_mean[7]`, `r priors$post_prior_sigma[7]`)), and higher covariate effect of burial values (N(`r priors$post_prior_mean[5]`, `r priors$post_prior_sigma[5]`)) to model an increased social differentiation.

@@

# Reproducibility and open source materials {.unnumbered}

The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at <https://doi.org/xxx/xxx> to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

# Results {.unnumbered}
